Appendix S1: Supplemental figures and table

### 2023-12-13 ms/ms.tex
preparing final submission

% toggle spacing
\usepackage{setspace}
% \singlespacing
\doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

\noindent \textbf{Premise of the study} The adaptive significance of
stomata on both upper and lower leaf surfaces, called amphistomy, is
unresolved. A widespread association between amphistomy and open, sunny
habitats suggests the adaptive benefit of amphistomy may be greatest in
these contexts, but this hypothesis has not been tested experimentally.
Understanding why amphistomy evolves can inform its potential as a
target for crop improvement and paleoenvironment reconstruction.

\textbf{Methods} We developed a new method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the log-ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf but with gas exchange
blocked through the upper (adaxial) surface, which we term
``pseudohypostomy''. We used humidity to modulate stomatal conductance
and thus compare photosynthetic rates at the same total stomatal
conductance. We estimated \(\mathrm{AA}\) and related physiological and
anatomical traits in 12 populations, six coastal (open, sunny) and six
montane (closed, shaded), of the indigenous Hawaiian species ʻilima
(\emph{Sida fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4.04 times more from
amphistomy compared to their montane counterparts. Our evidence was
equivocal with respect to two hypotheses -- that coastal leaves benefit
more because 1) they are thicker and therefore have lower CO\(_2\)
conductance through the internal airspace, and 2) that they benefit more
because they have similar conductance on each surface, as opposed to
most of the conductance being on the lower (abaxial) surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy \emph{per se} increases photosynthesis, consistent with the
hypothesis that parallel pathways through upper and lower mesophyll
increase the supply of CO\(_2\) to chloroplasts. The prevalence of
amphistomatous leaves in open, sunny habitats can partially be explained
the increased benefit of amphistomy in `sun' leaves, but the mechanistic
basis of this observation is an area for future research.

\textbf{Keywords:} amphstomy, leaf, light, Hawaiʻi, \emph{Sida fallax},
stomata

@@

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that may
help them to compete for light availability. This montane morphology is
not found in \emph{S. fallax} populations on other Pacific Islands
(Pejhanmehr, 2022).

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
NE, USA). We estimated the photosynthetic rate (\(A\)) and stomatal
conductance to water vapor (\(g_\text{sw}\)) at saturating light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). The midday
irradiance in coastal ʻilima typically meets or even exceeds a PPFD of
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) and previous
experiments with sun leaves revealed that
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) is always at or near
saturating irradiance. Even though lower irradiance may be saturating
for montane leaves, we used this higher value for all leaves to
standardize conditions.

@@

\autoref{fig:ags-curve} illustrates our method for collecting
\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
an amphi (untreated) curve and a pseudohypo (treated) curve. To control
for order effects, we alternated between starting with amphi or
pseudohypo leaf measurements, though we did not detect an effect of
treatment order on \(\mathrm{AA}\) (results not shown). In the field, we
acclimated the focal leaf to saturating light and high relative humidity
(\(\mathrm{RH} = 70\%\)), as described above, until \(A\) and
\(g_\text{sw}\) reach their maximum. We used these data as our estimates
of maximum \(A\) and \(g_\text{sw}\). After that, we decreased
\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
without biochemical downregulation. Hence, \(A_\text{amphi}\) and
\(A_\text{hypo}\) were both measured at low chamber humidity after the
leaf had acclimated to high humidity. All other environmental conditions
in the leaf chamber remained the same. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process of
acclimating the leaf to 70\% \(\mathrm{RH}\) and inducing stomatal
closure with low \(\mathrm{RH}\) with the other treatment (amphi or
pseudohypo).

@@

In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
those cases, we estimated \(\hat{A}_\text{amphi}\) and
\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
the range of overlap between the curves. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. The vertical lines in \autoref{fig:licor} show the
\(g_\text{sw}\) for each leaf. We estimated \(\mathrm{AA}\) from
\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) for each leaf using
the log-response ratio shown above.

To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
curve for each leaf, we fit Bayesian regressions using the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) with MCMC sampling
in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated parameters and confidence
intervals as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
95\% confidence intervals of
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.1
and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar, 2023). We
sampled the posterior distribution from 4 chains with 1000 iterations
each after 1000 warmup iterations per chain. We estimated the
relationship between population average leaf thickness and
\(\mathrm{AA}\) measured from a single individual per population. We
used this approach because most of the variation in leaf thickness was
among sites and the plant selected for gas exchange measurements was not
always among the plants randomly selected for leaf thickness, precluding
individual level correlation. We propagated uncertainty about in
\(\mathrm{AA}\) and leaf thickness estimates by integrating over the
entire posterior distribution sample for each variable. The key
prediction is that the effect of leaf thickness on \(\mathrm{AA}\) is
positive, meaning the 95\% confidence interval of the slope should be
positive and not encompass 0.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry and Češnovar,
2023). We interpolated missing adaxial guard cell lengths from 6 out of
185 samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations per chain.
From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
as:

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see \autoref{fig:licor} for individual
curves). Overall, the average \(\mathrm{AA}\) among coastal and montane
leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
\(0.027~[\numrange{-0.0034}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.039}{0.14}]\).
Posterior predictions closely match observed values of \(A\)
(\autoref{fig:pp-licor}), indicating an adequate model fit from which we
can interpolate between measurements reliably. It also suggests that
slight extrapolation beyond the data should be reliable, but this is
less certain. When we remove two leaves where we extrapolated slightly
beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
\(\mathrm{AA}_\text{coastal}\) is still positive,
\(0.081~[\numrange{0.023}{0.13}]\), but the difference between coastal
and montane leaves is smaller, \(0.053~[\numrange{-0.012}{0.12}]\), and
confidence intervals slightly overlap 0. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(\autoref{fig:habitat-Ags}). Water-use efficiency
(\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
(Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).

\begin{figure}[H]
  \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more from amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}

@@

Coastal and montane leaves have similar average
\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
water vapor (\autoref{fig:habitat-gmaxratio}); coastal leaves have
\(0.059~[\numrange{-0.14}{0.28}]\) higher \(g_\mathrm{smax,ratio}\) than
montane leaves, but the 95\% confidence intervals overlap 0 difference.
The \(g_\mathrm{smax,ratio}\) is somewhat bimodal among sites. Some
sites in both habitats have leaves with \(g_\mathrm{smax,ratio} < 0.07\)
and others with \(g_\mathrm{smax,ratio} > 0.2\)
(\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
montane sites where those on the Big Island of Hawaiʻi all have low
\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
\(g_\mathrm{smax,ratio}\). There is no relationship between
\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
(\autoref{fig:traits-aa}B;
\(\text{slope}~= 0.14~[\numrange{-0.057}{0.34}]\)) in our sample.

\begin{figure}[H]
  \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}

@@

Custom scripts are available on a GitHub repository
(https://github.com/cdmuir/stomata-ilima) and will be archived on Zenodo
with a DOI and stable URL upon publication. Raw data will be deposited
on Dryad with a DOI and stable URL upon publication. {[}THE GITHUB REPO
AND DRYAD DATA ARE AVAILABLE TO REVIEWERS{]}

@@

\hypertarget{literature-cited}{%
\section*{Literature Cited ---}\label{literature-cited}}
\addcontentsline{toc}{section}{Literature Cited ---}

@@

\leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
Gabry, J., and R. Češnovar. 2023.
\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
{R} {Interface} to '{CmdStan}'}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_phylogenetic_2022}{}}%
Pejhanmehr, M. 2022. Phylogenetic affinities among
{\textless{}}i{\textgreater{}}{Sida}{\textless{}}{\textbackslash{}}i{\textgreater{}}
species and allied genera ({Malvaceae}: {Malveae}), and examination of
{\textless{}}i{\textgreater{}}{Sida}
fallax{\textless{}}{\textbackslash{}}i{\textgreater{}} within the
{Hawaiian} {Islands} and throughout the {Pacific}. PhD thesis.
University of Hawai'i at Mānoa.

@@

>>>>

% toggle spacing
\usepackage{setspace}
\singlespacing
% \doublespacing

@@

\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, frame hidden]}{\end{tcolorbox}}\fi

@@

\noindent \textbf{Premise of the study} The adaptive significance of
stomata on both upper and lower leaf surfaces, called amphistomy, is
unresolved. A widespread association between amphistomy and open, sunny
habitats suggests the adaptive benefit of amphistomy may be greatest in
these contexts, but this hypothesis has not been tested experimentally.
Understanding amphistomy informs its potential as a target for crop
improvement and paleoenvironment reconstruction.

\textbf{Methods} We developed a method to quantify ``amphistomy
advantage'', \(\mathrm{AA}\), as the log-ratio of photosynthesis in an
amphistomatous leaf to that of the same leaf but with gas exchange
blocked through the upper surface (pseudohypostomy). Humidity modulated
stomatal conductance and thus enabled comparing photosynthesis at the
same total stomatal conductance. We estimated \(\mathrm{AA}\) and leaf
traits in six coastal (open, sunny) and six montane (closed, shaded)
populations of the indigenous Hawaiian species ʻilima (\emph{Sida
fallax}).

\textbf{Key results} Coastal ʻilima leaves benefit 4.04 times more from
amphistomy than montane leaves. Evidence was equivocal with respect to
two hypotheses -- that coastal leaves benefit more because 1) they are
thicker and have lower CO\(_2\) conductance through the internal
airspace, and 2) that they benefit more because they have similar
conductance on each surface, as opposed to most conductance being
through the lower surface.

\textbf{Conclusions} This is the first direct experimental evidence that
amphistomy increases photosynthesis, consistent with the hypothesis that
parallel pathways through upper and lower mesophyll increase CO\(_2\)
supply to chloroplasts. The prevalence of amphistomatous leaves in open,
sunny habitats can partially be explained the increased benefit of
amphistomy in `sun' leaves, but the mechanistic basis remains uncertain.

\textbf{Keywords:} amphistomy, Hawaiʻi, leaf, light, Malvaceae,
photosynthesis, \emph{Sida fallax}, stomata

@@

Here we focus on variation within an indigenous plant species \emph{Sida
fallax} Walp. (Malvaceae), known in the Hawaiian language as ʻilima.
ʻIlima is found from sea level to elevations \(>1000\) mas on multiple
Hawaiian islands. Coastal populations are morphologically different from
montane populations (\autoref{fig:study-system}). Coastal regions of
Hawaiʻi are characterized by high sun exposure, warmer temperatures,
high winds, salinity, and variation in water availability. Coastal
populations of ʻilima tend to be short and prostrate which likely helps
them to withstand the windy environment (\autoref{fig:study-system}a).
The leaves of these populations are covered on both surfaces in dense,
soft hairs that give the leaves a silvery green appearance
(\autoref{fig:study-system}b), which helps mitigate water loss by
reflecting solar radiation, thereby lowering leaf temperature
(Ehleringer and Björkman, 1978). Montane regions, on the other hand,
provide very different challenges. Many other tall species grow on the
slopes of these wet mountainous regions, which makes light competition a
factor that plants may need to adapt to. Possibly due to this, montane
populations are erect and shrub- or tree-like, capable of growing meters
tall with strong, woody stems. These individuals have smooth, green
foliage with serrated edges. Montane populations exhibit traits that may
help them to compete for light availability. This montane morphology is
not found in \emph{S. fallax} populations on other Pacific Islands
(Pejhanmehr et al., 2023).

@@

At each site, we selected one representative leaf from one plant near
the middle of the transect for gas exchange measurements using a
portable infrared gas analyzer (LI-6800PF, LI-COR Biosciences, Lincoln,
Nebraska, USA). We estimated the photosynthetic rate (\(A\)) and
stomatal conductance to water vapor (\(g_\text{sw}\)) at saturating
light
(\(\text{photosynthetic photon flux density (PPFD)} = 2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)),
ambient CO\(_2\) (415 ppm), and
\(T_\mathrm{leaf} = 25.0 \textendash 29.3° \text{C}\). The midday
irradiance in coastal ʻilima typically meets or even exceeds a PPFD of
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) and previous
experiments with sun leaves revealed that
\(2000~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\) is always at or near
saturating irradiance. Even though lower irradiance may be saturating
for montane leaves, we used this higher value for all leaves to
standardize conditions.

@@

Appendix S1 (see the Supplementary Data with this article),
\autoref{fig:ags-curve} illustrates our method for collecting
\(A \textendash g_\text{sw}\) curves. We collected two curves per leaf,
an amphi (untreated) curve and a pseudohypo (treated) curve. To control
for order effects, we alternated between starting with amphi or
pseudohypo leaf measurements, though we did not detect an effect of
treatment order on \(\mathrm{AA}\) (results not shown). In the field, we
acclimated the focal leaf to saturating light and high relative humidity
(\(\mathrm{RH} = 70\%\)), as described above, until \(A\) and
\(g_\text{sw}\) reach their maximum. We used these data as our estimates
of maximum \(A\) and \(g_\text{sw}\). After that, we decreased
\(\mathrm{RH}\) to \(\approx 10\%\) to induce rapid stomatal closure
without biochemical downregulation. Hence, \(A_\text{amphi}\) and
\(A_\text{hypo}\) were both measured at low chamber humidity after the
leaf had acclimated to high humidity. All other environmental conditions
in the leaf chamber remained the same. We logged data until
\(g_\text{sw}\) reached its nadir. We then repeated the process of
acclimating the leaf to 70\% \(\mathrm{RH}\) and inducing stomatal
closure with low \(\mathrm{RH}\) with the other treatment (amphi or
pseudohypo).

@@

In 10 of 12 leaves, the minimum \(g_\text{sw}\) of the amphi curve was
smaller than the maximum \(g_\text{sw}\) of the pseudohypo curve
(i.e.~the curves overlapped for a range of \(g_\text{sw}\) values). In
those cases, we estimated \(\hat{A}_\text{amphi}\) and
\(\hat{A}_\text{hypo}\) at the \(g_\mathrm{sw}\) value in the middle of
the range of overlap between the curves. In 2 of 12 leaves, the
\(A \textendash g_\text{sw}\) curves did not quite overlap because the
minimum \(g_\text{sw}\) of the amphi curve was slightly greater than the
maximum \(g_\text{sw}\) of the pseudohypo curve. In those cases, we
estimated \(\mathrm{AA}\) by extrapolating slightly,
\(1.98\times 10^{-3}~\text{and}~3.29\times 10^{-3}~\text{mol}~\text{m}^{-2}~\text{s}^{-1}\),
beyond the measured curves to the \(g_\mathrm{sw}\) value in between the
curves. The vertical lines in Appendix S1, \autoref{fig:licor} show the
\(g_\text{sw}\) for each leaf. We estimated \(\mathrm{AA}\) from
\(\hat{A}_\text{amphi}\) and \(\hat{A}_\text{hypo}\) for each leaf using
the log-response ratio shown above.

To estimate \(\hat{\beta}\)'s from the \(A \textendash g_\text{sw}\)
curve for each leaf, we fit Bayesian regressions using the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) with MCMC sampling
in \emph{Stan} (Stan Development Team, 2023). We used CmdStan version
2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023) to
interface with \emph{R} version 4.3.1 (R Core Team, 2023). We sampled
the posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated parameters and confidence
intervals as the median and 95\% quantile intervals of the posterior,
respectively. The key prediction is that
\(\mathrm{AA}_\text{coastal} > \mathrm{AA}_\text{montane}\), meaning the
95\% confidence intervals of
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane}\) should be
positive and not encompass 0.

@@

We tested whether leaf thickness (log-transformed) varied between
coastal and montane populations and among individuals within populations
using a Bayesian mixed-effects model with habitat as a fixed effect and
individual plant and site as random effects. We used the \emph{R}
package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit the model in
\emph{Stan} (Stan Development Team, 2023) with CmdStan version 2.33.1
and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023). We sampled the
posterior distribution from 4 chains with 1000 iterations each after
1000 warmup iterations per chain. We estimated the relationship between
population average leaf thickness and \(\mathrm{AA}\) measured from a
single individual per population. We used this approach because most of
the variation in leaf thickness was among sites and the plant selected
for gas exchange measurements was not always among the plants randomly
selected for leaf thickness, precluding individual level correlation. We
propagated uncertainty about in \(\mathrm{AA}\) and leaf thickness
estimates by integrating over the entire posterior distribution sample
for each variable. The key prediction is that the effect of leaf
thickness on \(\mathrm{AA}\) is positive, meaning the 95\% confidence
interval of the slope should be positive and not encompass 0.

@@

We tested whether \(g_\mathrm{smax,ratio}\) varied between coastal and
montane populations and among individuals within populations using a
Bayesian mutliresponse, mixed-effects model. The modeled response
variables are stomatal count and guard cell length on each surface.
Counts were modeled as negative binomially distributed variable from a
latent stomatal density and a parameter \(\phi\) to estimate
overdispersion in counts relative to a Poisson model. For all traits,
the explanatory variables were habitat as a fixed effect and leaf within
individual plant, individual plant, and site as random effects. We used
the \emph{R} package \textbf{brms} version 2.20.4 (Bürkner, 2017) to fit
the model in \emph{Stan} (Stan Development Team, 2023) with CmdStan
version 2.33.1 and \textbf{cmdstanr} version 0.6.1 (Gabry et al., 2023).
We interpolated missing adaxial guard cell lengths from 6 out of 185
samples with zero adaxial stomata using the ``mi'' function in
\textbf{brms} package. We sampled the posterior distribution from 4
chains with 1000 iterations each after 1000 warmup iterations per chain.
From each posterior sample, we calculated \(g_{\mathrm{smax,ratio}}\)
as:

@@

Amphistomy increases photosynthesis in leaves of coastal ʻilima plants
more than that of montane plants. \(\mathrm{AA}\) was significantly
greater than 0 (95\% confidence intervals did not overlap 0) in 5 of 6
coastal leaves, but only 1 of 6 montane leaves
(\autoref{fig:habitat-aa}; see Appendix S1, \autoref{fig:licor} for
individual curves). Overall, the average \(\mathrm{AA}\) among coastal
and montane leaves is \(0.12~[\numrange{0.077}{0.15}]\) and
\(0.027~[\numrange{-0.0034}{0.057}]\), respectively; the difference in
average \(\mathrm{AA}\) between habitat types is
\(\mathrm{AA}_\text{coastal} - \mathrm{AA}_\text{montane} = 0.09~[\numrange{0.039}{0.14}]\).
Posterior predictions closely match observed values of \(A\) (Appendix
S1, \autoref{fig:pp-licor}), indicating an adequate model fit from which
we can interpolate between measurements reliably. It also suggests that
slight extrapolation beyond the data should be reliable, but this is
less certain. When we remove two leaves where we extrapolated slightly
beyond fitted \(A \textendash g_\mathrm{sw}\) curves, we estimate that
\(\mathrm{AA}_\text{coastal}\) is still positive,
\(0.081~[\numrange{0.023}{0.13}]\), but the difference between coastal
and montane leaves is smaller, \(0.053~[\numrange{-0.012}{0.12}]\), and
confidence intervals slightly overlap 0. Maximum photosynthetic rate was
slightly, but not significantly higher in coastal leaves (Welch Two
Sample \emph{t}-test, \(t_{9.65} = 1.6\), \(P = 0.14\)); total stomatal
conductance was similar (Welch Two Sample \emph{t}-test,
\(t_{9.71} = -0.09\), \(P = 0.93\)) in coastal and montane leaves
(Appendix S1, \autoref{fig:habitat-Ags}). Water-use efficiency
(\(A / g_\mathrm{sw}\)) was significantly higher in coastal leaves
(Welch Two Sample \emph{t}-test, \(t_{9.99} = 2.54\), \(P = 0.03\)).

@@

Coastal and montane leaves have similar average
\(g_\mathrm{smax,ratio}\), the ratio of adaxial (upper) to total
\(g_\mathrm{smax}\), the anatomical maximum stomatal conductance to
water vapor (Appendix S1, \autoref{fig:habitat-gmaxratio}); coastal
leaves have \(0.059~[\numrange{-0.14}{0.28}]\) higher
\(g_\mathrm{smax,ratio}\) than montane leaves, but the 95\% confidence
intervals overlap 0 difference. The \(g_\mathrm{smax,ratio}\) is
somewhat bimodal among sites. Some sites in both habitats have leaves
with \(g_\mathrm{smax,ratio} < 0.07\) and others with
\(g_\mathrm{smax,ratio} > 0.2\) (Appendix S1,
\autoref{fig:habitat-gmaxratio}). This is particularly noticeable in
montane sites where those on the Big Island of Hawaiʻi all have low
\(g_\mathrm{smax,ratio}\) whereas those on Oʻahu have relatively high
\(g_\mathrm{smax,ratio}\). There is no relationship between
\(g_\mathrm{smax,ratio}\) and \(\mathrm{AA}\) in either habitat
(\autoref{fig:traits-aa}B;
\(\text{slope}~= 0.14~[\numrange{-0.057}{0.34}]\)) in our sample.

@@

Custom scripts are available on a GitHub repository
(https://github.com/cdmuir/stomata-ilima) and archived on Zenodo (). Raw
data are deposited on Dryad (https://doi.org/10.5061/dryad.rxwdbrvfw ).

@@

\hypertarget{literature-cited}{%
\section{Literature Cited ---}\label{literature-cited}}

@@

\leavevmode\vadjust pre{\hypertarget{ref-gabry_cmdstanr_2023}{}}%
Gabry, J., R. Češnovar, and A. Johnson. 2023.
\href{https://mc-stan.org/cmdstanr,\%20https://discourse.mc-stan.org}{Cmdstanr:
{R} {Interface} to '{CmdStan}'}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-pejhanmehr_low_2023}{}}%
Pejhanmehr, M., M. Yorkston, and C. W. Morden. 2023.
\href{https://doi.org/10.2984/76.4.2}{Low {Genetic} {Diversity} in the
{Highly} {Morphologically} {Diverse} {Sida} fallax {Walp}. ({Malvaceae})
{Throughout} the {Pacific1}}. \emph{Pacific Science} 76.

@@

\clearpage

\hypertarget{figure-legends}{%
\section{Figure Legends ---}\label{figure-legends}}

\begin{figure}[H]
  % \includegraphics[height=3.25in]{../figures/study-system.pdf}
  % \includegraphics[height=3.25in]{../figures/habitat-climate.pdf}
  \caption{A. Typical growth form of montane (left) and coastal (right) ‘ilima plants and B. leaves. C. Map of the sites that were sampled on the islands of Oʻahu and Hawaiʻi (aka Big Island). D. Climatic, light, and vegetation height comparisons between montane (blue) and coastal (orange) habitats sampled in this study. Open circles are values for the midpoint of each site transect; closed circles and intervals are the mean $\pm~1$ standard error. The habitats differ significantly in mean annual precipitation (top-left), solar radiation (bottom-left), temperature (top-right), and vegetation height (bottom-right). MAP = mean annual precipitation; MAT = mean annual temperature; ns = not significant; * indicates $0.01 \le P < 0.05$; ** indicates $0.001 \le P < 0.01$.}
  \label{fig:study-system}
\end{figure}

\begin{figure}[H]
  % \includegraphics{../figures/habitat-aa.pdf}
  \caption{Coastal leaves benefit more from amphistomy than montane leaves. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of amphistomy advantage for that leaf. Each leaf is from a different montane (blue) or coastal (orange) site, arranged by habitat and ascending amphistomy advantage within habitat. The longer horizonal bars are the average amphistomy advantage for montane and coastal leaves. $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:habitat-aa}
\end{figure}

\begin{figure}[H]
  % \includegraphics{../figures/traits-aa.pdf}
  \caption{Relationships between leaf amphistomy advantage, (A) leaf thickness and (B) $g_\mathrm{smax,ratio}$ among ʻilima (\textit{Sida fallax}) plants from montane (blue) and coastal (orange) habitats in Hawaiʻi. A positive amphistomy advantage ($y$-axis) means that the photosynthetic rate of an amphistomatous leaf is greater than that of an identical pseudohypostomatous leaf at the same overall $g_\mathrm{sw}$. Each point-interval is the median posterior estimate plus 95\% confidence interval of the trait value. Each leaf is from a different montane (blue) or coastal (orange) site. Lines are the estimated linear regression of (A) log(leaf thickness) and (B) $g_\mathrm{smax,ratio}$ on amphistomy advantage; ribbons are the 95\% confident bands of the regression. Symbols: $g_\mathrm{smax,ratio}$, anatomical maximum stomatal conductance ratio; $g_\mathrm{sw}$, stomatal conductance to water vapor.}
  \label{fig:traits-aa}
\end{figure}

### 2023-12-13 ms/si.qmd
preparing final submission

---
format:
  pdf:
    keep-tex: true  
    include-in-header: 
      - preamble.tex
    toc: false
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2023 – Appendix S1}
mainfont: Times New Roman
---

>>>>

---
title: "Appendix S1: Supplemental figures and table"
format:
  pdf:
    keep-tex: true  
    include-in-header: 
      - preamble.tex
    toc: false
header-includes: |
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhead[C]{Triplett \textit{et al.} — \textit{American Journal of Botany} 2024 – Appendix S1}
  \addtokomafont{disposition}{\rmfamily}
  \RedeclareSectionCommand[
    font=\normalfont\Large]{section}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\bfseries]{subsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize\itshape]{subsubsection}
  \RedeclareSectionCommand[
    font=\normalfont\normalsize]{paragraph}
mainfont: Times New Roman
---
